---
title: "Mapping_Basics_World"
author: "Julia Christensen"
date: "March 25, 2019"
output: html_document
---


****************************************************
# 1) SET UP 

## 1.1. General Set Up
```{r}
### Clear global environment
rm(list=ls()) 
```

## 1.2. Libraries 
```{r}
### Download Libraries 
library(pacman)
p_load(tidyverse, data.table, rmarkdown, knitr, tinytex, magrittr,
       gplots, #for heatmaps
       viridis, RColorBrewer, scales, #good colors
       grDevices, graphics,
       mapproj, ggmap,
       rgdal, tmap, 
       maptools, tmaptools,
       wbstats, WDI, #use to download world bank data
       sf, rnaturalearth
       )
```
Notes:
*Not all of these packages are used in this file. In general, it's not a good idea to have a bunch of extra packages (sometimes two packages have a function with the same name, which causes chaos). However, I don't remember which ones are necessary.


****************************************************
# 2) DOWNLOAD SHAPEFILE

## 2.1. Use "rnaturalearth" to download shapefile for world 
```{r}
world_geo <- rnaturalearth::ne_countries(scale = 50, returnclass = "sp")
```

## 2.2. Save tidied shapefile  
```{r, message = FALSE, warning = FALSE}
# Create map dataframe
world_geo_tidy <- data.table(broom::tidy(world_geo)) 

# Look at map dataframe (Note: no FIPS column)
world_geo_tidy %>% head()
```
Note(s):
*In general, it's bad practice to turn off warnings for a chunk, but I've done so here because printing the warnings makes R take (a lot) longer. This warning is only about coercing factors and characters. 
*For chunk options, see https://rmarkdown.rstudio.com/lesson-3.html

## 2.3. Crosswalk for census id and FIPS 
```{r}
# Save data in data slot
crosswalk <- world_geo@data

# Save row IDs in a column 
crosswalk <- cbind(id=rownames(crosswalk),crosswalk)
```


****************************************************
# 3) DOWNLOAD WORLD BANK DATA

## 3.1. Select variables to download 
```{r}
# Set Variables 
selectvars <- c(
  # Maternal mortality ratio (modeled estimate, per 100,000 live births)
  "SH.STA.MMRT", 
  # Births attended by skilled health staff (% of total)
  "SH.STA.BRTC.ZS", 
  # Sex ratio at birth (male births per female births)
  "SP.POP.BRTH.MF", 
  # Adolescent fertility rate (births per 1,000 women ages 15-19)
  "SP.ADO.TFRT", 
  # Mortality rate, under-5 (per 1,000 live births)
  "SH.DYN.MORT", 
  # Mortality rate, infant (per 1,000 live births)
  "SP.DYN.IMRT.IN", 
  # School enrollment, primary (gross), gender parity index (GPI)
  "SE.ENR.PRIM.FM.ZS", 
  # School enrollment, secondary (gross), gender parity index (GPI)
  "SE.ENR.SECO.FM.ZS", 
  # School enrollment, tertiary (gross), gender parity index (GPI)
  "SE.ENR.TERT.FM.ZS", 
  # Pregnant women receiving prenatal care (%)
  "SH.STA.ANVC.ZS", 
  # Women who were first married by age 15 (% of women ages 20-24
  "SP.M15.2024.FE.ZS", 
  # Female genital mutilation prevalence (%)
  "SH.STA.FGMS.ZS", 
  # Women making their own informed decisions regarding sexual relations, contraceptive use and a
  # reproductive health care  (% of women age 15-49)
  "SG.DMK.SRCR.FN.ZS", 
  # Women who believe a husband is justified in beating his wife when she argues with him (%)
  "SG.VAW.ARGU.ZS")
```

## 3.2. Two packages
```{r}
?wb
?WDI
```

wb() - This function downloads the requested information using the World Bank API

WDI() - Downloads the requested data by using the World Bank's API, parses the resulting XML file, and formats it in long country-year format.

## 3.3. Download data
```{r}
# Use wbstats
wb_wb <- wbstats::wb(country = "countries_only", 
                     indicator = selectvars, 
                     start=2010,
                     end=2015,
                     return_wide=TRUE)
# Use WDI
wb_WDI <- WDI::WDI(country="all", 
                   indicator = selectvars, 
                   start=2010,
                   end=2015,
                   extra=TRUE #returns extra variables such as region, iso3c code, and incomeLevel
                   )
```

## 3.3. Compare Packages

wb() dropped one variable (SP.M15.2024.FE.ZS), but overall, seems like a better package. It has more options, and better country subsetting abilities. 
```{r}
wb_wb %>% glimpse()
```

WDI() is more limited overall, but successfully downloaded all variables. 
```{r}
wb_WDI %>% glimpse()
```

## 3.4. Select Package Results
```{r}
# Rename selected results
wb_dat <- wb_wb

# Drop other results
rm(wb_wb, wb_WDI)
```


****************************************************
# 4) MERGE

# 4.1. Merge WB with Croswalk
```{r}
# Subset crosswalk to only needed variables
crosswalk %<>% select(id, iso_a2)

# Use crosswalk to merge race data with shapefile IDs
wb_wID <- left_join(x = crosswalk, 
                    y = wb_dat, 
                    by = c("iso_a2" = "iso2c"))
```

# 4.1. Merge WB with Shapefile
```{r}
# Merge census data with map dataframe 
map_df <- left_join(x = world_geo_tidy, 
                    y = wb_wID, 
                    by = "id")
```

# 5) GENERATE MAPS

## 5.1. Rename vars
```{r}
map_df %<>%
  # Rename variables 
    rename("Sex_ratio"="SP.POP.BRTH.MF",
           "Prenatal"="SH.STA.ANVC.ZS",
           "Child_marriage" ="SP.M15.2024.FE.ZS",
           "Infant_mortality"="SP.DYN.IMRT.IN") 
```

## 5.2. Plot one variable
```{r}
# Generate map of 2010 data 
map_df %>%
  filter(date == 2015) %>%
  ggplot(aes(x=long, y=lat, group=group)) +
  # Plot data for each county (based on fill_var)
    geom_polygon(aes_string(fill = "Infant_mortality"), colour="white", size=.1) +
  # See note below (related to how 2D map created for 3D globe)
    coord_quickmap() +
  # Titles / labels (based on txt_var)  
    labs(title=paste0("Infant Mortality"), x="", y="") +
  # Theme 
    theme_void() +
  # Fill
    scale_fill_viridis(alpha = .75, begin = 0, end = 1, 
                       direction = -1, discrete = FALSE, option = "C",
                       #trans=log10_trans(),
                       name = "Key:")  
```
