---
title: "Data visualization basics"
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
rm(list=ls())

# install packages 

pacman::p_load(
  ggplot2, 
  tidyverse,
  ggthemes,
  ggrepel,
  broom
)

```

# Motivation 

- The following material is adapted from Kieran Healy's wonderful book (2018) on [data visualization](https://socviz.co/) and Hadley Wickham's equally wonderful book on [ggplot2](https://link.springer.com/content/pdf/10.1007%2F978-0-387-98141-3.pdf). For more theoretical discussions, I recommend you to read [The Grammar of Graphics](https://link.springer.com/book/10.1007%2F0-387-28695-0) by Leland Wilkinson.

- Why should we care data visualization? More precisely, why should we learn the grammar of statistical graphics?
- Sometimes, pictures are better tools than words in 1) exploring, 2) understanding, and 3) explaining data.

## Anscombe's quartet

[Anscombe](https://en.wikipedia.org/wiki/Frank_Anscombe)'s quarter comprises four datasets, which are so alike in terms of their descriptive statistics but quite different when presented graphically.

```{r}

# data 
anscombe 

```

```{r}

# correlation 
cor(anscombe)[c(1:4),c(5:8)]

```

```{r}

# gather and select 
anscombe_processed <- anscombe %>%
  gather(x_name, x_value, x1:x4) %>%
  gather(y_name, y_value, y1:y4)
  
# plot
anscombe_processed %>%
  ggplot(aes(x = x_value, y = y_value)) +
    geom_point() +
    geom_smooth(method = lm, se = FALSE) +
    facet_grid(x_name ~ y_name) +
    theme_bw() +
    labs(x = "X values",
       y = "Y values",
       title = "Anscombe's quartet")
```

# ggplot2 basics 

- the grammar of graphics 

    - data
    - aesthetic attributes (color, shape, size)
    - geometric objects (points, lines, bars)
    - stats (summary stats)
    - scales (map values in he data space)
    - coord (data coordinates)
    - facet (facetting specifications)
    
No worries for new terms. We're going to learn them by actually plotting. 

- Workflow: 

    1. Tidy data 
    2. Mapping 
    3. Geom 
    4. Cor_ordinates and scales 
    5. Labels and guides
    6. Themes
    7. Save files 

## Tidy data 

- We covered tidy data in the previous sessions. 

```{r, echo=FALSE}
# load library 
library(gapminder)

gapminder
```

## Data

- Data argument tells where to find the variable you're about to use.

## Mapping and Geom

- aes (aesthetic mappings or aesthetics) tells which variables (x, y) in your data should be represented by which visual elements (color, shape, size) in the plot.

- geom_ tells the type of plot you are going to use 

### Basic aes (x , y)

```{r}

p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp)) # ggplot or R in general takes positional arguments too. So, you don't need to name data, mapping each time you use ggplot2.

str(p) # a list of data and functions, but it shows nothing 

p + geom_point()

p + geom_point() + geom_smooth() # geom_smooth has calculated a smoothed line;
                                 # the shaded area is the standard error for the line
```

### What if I am interested in plotting only one distribution (univariate distribution)

- `geom_histogram()`: For the probability distributoin of a continuous variable. Bins divide the entire range of values into a series of intervals (see [the Wiki entry](https://en.wikipedia.org/wiki/Histogram)). 
- `geom_density()`: Also for the proability distributoin of a continuous variable. It calculates a [kernel density estimate](https://en.wikipedia.org/wiki/Kernel_density_estimation) of the underlying distribution. 

#### Histogram 
```{r}

data(midwest) # load midwest dataset 

midwest
```

```{r, eval = FALSE}
midwest %>%
  ggplot(aes(x = area)) +
  geom_point() # not working. 
```

```{r}
midwest %>%
  ggplot(aes(x = area)) +
  geom_histogram() # stat_bin argument picks up 30 bins (or "bucket") by default.

midwest %>%
  ggplot(aes(x = area)) +
  geom_histogram(bins = 10) # only 10 bins. 

ggplot(data = subset(midwest, state %in% c("OH","IN")),
         mapping = aes(x = percollege, fill = state)) +
  geom_histogram(alpha = 0.4, bins = 20)


```

#### Density 

```{r}
midwest %>%
  ggplot(aes(x = area, fill = state, color = state)) +
    geom_density(alpha = 0.3)
```

### Advanced aes (size, color)

- There's also fill argument (mostly used in geom_bar()). Color aes affects the appearance of lines and points, fill is for the filled areas of bars, polygons, and in some cases, the interior of a smoother's standard error ribon.

- The property size/color/fill represents... 

```{r}
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          size = pop)) +
  geom_point()
```

```{r}
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          size = pop,
                          color = continent)) +
  geom_point()

# try red instead of "red"
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          size = pop,
                          color = "red")) + 
  geom_point()
```

Aesthetics also can be mapped per Geom. 

```{r}
p + geom_point() + 
  geom_smooth()

p + geom_point(alpha = 0.3) + # alpha controls transparency
  geom_smooth(color = "red", se = FALSE, size = 2)

p + geom_point(alpha = 0.3) + # alpha controls transparency
  geom_smooth(color = "red", se = FALSE, size = 2, method = "lm")
```

```{r}
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          color = continent)) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder")

ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          color = continent,
                          fill = continent)) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder")
```

## Co-ordinates and scales 

```{r}

p + geom_point() +
  coord_flip() # coord_type
```

The data is heavily bunched up against the left side. 
```{r}

p + geom_point() # without scaling 

p + geom_point() +
  scale_x_log10() # scales the axis of a plot to a log 10 basis 

p + geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10()

p + geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10()
```

```{r}
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          color = continent)) + 
  geom_point(alpha = 0.3) +
  scale_x_log10() + 
  geom_smooth(method = "loess", color = "red") +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder")

ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          color = continent,
                          fill = continent)) + 
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  geom_smooth(method = "loess", color = "red") +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder")
```

## Labels and guides 

`scales` package has some useful premade formatting functions. You can either load scales or just grab the function you need from the library using `scales::` 

```{r}
p + geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder")
```

6. Themes
```{r}
p + geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder") +
  theme_economist()
```

## Save files 

- I highly recommend to save your file in a subdirectory named output or figures or something like that.

```{r}
figure_example <- p + geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder") +
  theme_economist()

ggsave(filename = "./figures/figure_example.png", plot = figure_example)
```