---
title: "Data visualization basics"
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
rm(list=ls())

# install packages 

pacman::p_load(
  ggplot2, 
  tidyverse,
  ggthemes,
  ggrepel,
  broom
)

```

# Motivation 

- The following material is adapted from Kieran Healy's wonderful book (2018) on [data visualization](https://socviz.co/). 
- Why should we care?
- Sometimes, pictures are better tools than words in 1) exploring, 2) understanding, and 3) explaining data.

## Anscombe's quartet

[Anscombe](https://en.wikipedia.org/wiki/Frank_Anscombe)'s quarter comprises four datasets, which are so alike in terms of their descriptive statistics but quite different when presented graphically.
```{r}

# data 
anscombe 

```

```{r}

# correlation 
cor(anscombe)[c(1:4),c(5:8)]

```

```{r}

# plot
anscombe %>%
  gather(x_name, x_value, x1:x4) %>%
  gather(y_name, y_value, y1:y4) %>%
  ggplot(aes(x = x_value, y = y_value)) +
           geom_point() +
           facet_grid(x_name ~ y_name) +
  theme_bw() +
  labs(x = "X values",
       y = "Y values",
       title = "Anscombe's quartet")
```

# ggplot2 basics 

- Workflow: 
  1. Tidy data 
  2. Mapping 
  3. Geom 
  4. Cor_ordinates and scales 
  5. Labels and guides
  6. Themes
  7. Save files 

## Tidy data 

- We covered tiday data in the previous sessions. 

```{r, echo=FALSE}
# load library 
library(gapminder)

gapminder
```

What is the difference between typeof and class?

```{r}
typeof(gapminder)

class(gapminder)
```

## Mapping and Geom

- ggplot tells what is your data 

- aes (aesthetic mappings or aesthetics) tells what is your variables of interests in the data

- geom_ tlles the type of plot you are going to use 

### Basic aes (x , y)

```{r}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))

p + geom_point()

p + geom_point() + geom_smooth()
```

### Advanced aes (size, color)

- There's also fill argument (mostly used in geom_bar()). 

- The property size/color/fill represents... 

```{r}
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          size = pop)) +
  geom_point()
```

```{r}
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          size = pop,
                          color = continent)) +
  geom_point()

# try red instead of "red"
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          size = pop,
                          color = "red")) + 
  geom_point()
```

```{r}
p + geom_point() + 
  geom_smooth()

p + geom_point(alpha = 0.3) + # alpha controls transparency
  geom_smooth(color = "red", se = FALSE, size = 2)
```


## Co-ordinates and scales 

```{r}

p + geom_point() +
  coord_flip() # coord_type
```

```{r}

p + geom_point() +
  scale_x_log10() # scale_mapping_type

p + geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10()

p + geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10()
```

## Labels and guides 
```{r}
p + geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder")
```

```{r}
ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          color = continent)) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder")

ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          color = continent,
                          fill = continent)) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder")
```

6. Themes
```{r}
p + geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder") +
  theme_economist()
```

## Save files 

- I highly recommend to save your file in a subdirectory named output or figures or something like that.

```{r}
figure_example <- p + geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", color = "red") +
  scale_x_log10(labels = scales::dollar) +
  labs( x = "log GDP",
        y = "Life Expectancy",
        title = "A Gapminder Plot",
        subtitle = "Data points are country-years",
        caption = "Source: Gapminder") +
  theme_economist()

ggsave(filename = "figure_example.png", plot = figure_example)

ggsave(filename = "figure_example_modified.png", plot = figure_example,
       height = 8,
       width = 10,
       units = "in")

```