---
title: "Data visualization intermdiates"
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
rm(list=ls())

# install packages 

pacman::p_load(
  ggplot2, 
  tidyverse,
  ggthemes,
  ggrepel,
  broom,
  gapminder
)

```

# Motivation 

- The following material is adapted from Kieran Healy's wonderful book (2018) on [data visualization](https://socviz.co/) and Hadley Wickham's equally wonderful on [ggplot2](https://link.springer.com/content/pdf/10.1007%2F978-0-387-98141-3.pdf). 

# ggplot2 intermediates 

## Grouping and facetting 

- Can you guess what's wrong?

```{r}

p <- ggplot(gapminder, aes(x = year, y = gdpPercap)) 

p + geom_point()

p + geom_line() 
```

```{r}
gapminder
```

- Use grouping and facetting to clarify 

```{r}
p <- ggplot(gapminder, aes(x = year, y = gdpPercap)) 

p + geom_line(aes(group = country)) # group by 

p + geom_line(aes(group = country)) + facet_wrap(~continent) # facetting 

p + geom_line(aes(group = country), color = "gray70") +
  geom_smooth(size = 1.1, method = "loess", se = FALSE) +
  scale_y_log10(labels = scales::dollar) +
  facet_wrap(~continent, ncol = 5) + # for single categorical variable; for multiple categorical variables use facet_grid()
  labs(x = "Year",
       y = "GDP per capita",
       title = "GDP per capita on Five continents") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Use pipes to summarize data

```{r}

gapminder %>%
  group_by(continent, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = lifeExp_mean, color = continent)) +
  geom_point() +
    labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy on Five continents")

gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = lifeExp_mean, color = country)) +
  geom_point() +
    labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in Europe")

gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = lifeExp_mean)) +
  geom_point() +
    labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  facet_wrap(~country)

```

```{r}
gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = country, y = lifeExp_mean)) +
  geom_boxplot() +
    labs(x = "Country",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  coord_flip()

gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = reorder(country, lifeExp_mean), y = lifeExp_mean)) +
  geom_boxplot() +
    labs(x = "Country",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  coord_flip()

gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = reorder(country, -lifeExp_mean), y = lifeExp_mean)) +
  geom_boxplot() +
    labs(x = "Country",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  coord_flip()
```

## Plotting text

```{r}
gapminder %>%
  filter(continent == "Asia" | continent == "Americas") %>%
  group_by(continent, country) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = gdp_mean, y = lifeExp_mean)) +
  geom_point() +
  geom_text(aes(label = country)) +
  scale_x_log10() +
  facet_grid(~continent)
```

## Ploting models 

In plotting models, we use David Robinson's [broom package](https://cran.r-project.org/web/packages/broom/vignettes/broom.html) in R extensively. The idea is to transform model outputs (i.e., predictions and estimations) into tidy objects so that we can combine, separate, and visualize these elements easily. 

```{r}

# regression model
out <-lm(formula = lifeExp ~ gdpPercap + pop + continent, 
          data = gapminder)
```

Tidy is a method in broom package. It "constructs a dataframe that summarizes the model's statistical findings". As the description states, tidy is a function that can be used generally for various models. For instance, a tidy can extract following information from a regression model.

- Term: a term being estimated 
- p.value
- statistic: a test statistic used to compute p-value
- estimate 
- conf.low: the low end of a confidence interval 
- conf.high: the high end of a confidence interval
- df: degrees of freedom

**Challege**

Try glance(out), what did you get from these commands? If you're curious, you can try ?glance.

```{r}
# estimates 
out_comp <- tidy(out)

p <- out_comp %>%
  ggplot(aes(x = term, y = estimate))

p + geom_point() +
  coord_flip() +
  theme_bw()
```

```{r}
# plus confidence intervals 
out_conf <- tidy(out, conf.int = TRUE)

# plotting coefficients using ggplot2 (pointrange)
out_conf %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() + coord_flip() + labs(x = "", y = "OLS Estimate") +
  theme_bw()

# another way to do it (errorbar)
out_conf %>%
  ggplot(aes(x = estimate, y = reorder(term, estimate))) +
  geom_point() + 
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  labs(y = "", x = "OLS Estimate") +
  theme_bw()
```

**Challenge**

1. If we only want to visualize a certain subset of variables, let's say gdpPercap and pop, how can you do that? Also, gdpPercap might be not very informative. What's the best way to change the value name?

2. broom is a great package for running split-and-combine regressions. See the following example and write down your workflow for visualize it.


```{r}
gapminder %>%
  group_by(continent) %>%
    do(tidy(lm(gdpPercap ~ lifeExp, data = .), conf.int = TRUE))

```

