---
title: "Data visualization intermdiates"
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
rm(list=ls())

# install packages 

pacman::p_load(
  ggplot2, 
  tidyverse,
  ggthemes,
  ggrepel,
  broom,
  gapminder
)

```

# Motivation 

- The following material is adapted from Kieran Healy's wonderful book (2018) on [data visualization](https://socviz.co/) and Hadley Wickham's equally wonderful on [ggplot2](https://link.springer.com/content/pdf/10.1007%2F978-0-387-98141-3.pdf). 

# ggplot2 intermediates 

Basic ideas:

- Grouping: tell `ggplot2` about the structure of your data 
- Facetting: break up your data into pieces for a plot 
- Transforming: perform some calculations on or summarize your data before producing the plot 

## Grouping

- Can you guess what's wrong?

```{r}

p <- ggplot(gapminder, aes(x = year, y = gdpPercap)) 

p + geom_point()

p + geom_line() 
```

`geom_line` joins up all the lines for each particular year in the order they appear in the dataset. `ggplot2` does not know the yearly observations in your data are grouped by country. 

Note that you need grouping when the grouping information you need to tell is not built into the varaibles being mapped (like continent).

```{r}
gapminder
```

## Facetting 

Facetting is to make small multiples. 

- `facet_wrap`: based on a single categorical variable like `facet_wrap(~single_categorical_variable)`. Your panels will be laid out in order and then wrapped into a grid.

- `facet_grid`: when you want to cross-classify some data by two categorical variables like `facet_grid(one_cat_variable ~ two_cat_variable)`. 

```{r}
p <- ggplot(gapminder, aes(x = year, y = gdpPercap)) 

p + geom_line(aes(group = country)) # group by, # The outlier is Kuwait. 

p + geom_line(aes(group = country)) + facet_wrap(~continent) # facetting 

p + geom_line(aes(group = country), color = "gray70") +
  geom_smooth(size = 1.1, method = "loess", se = FALSE) +
  scale_y_log10(labels = scales::dollar) +
  facet_wrap(~continent, ncol = 5) + # for single categorical variable; for multiple categorical variables use facet_grid()
  labs(x = "Year",
       y = "GDP per capita",
       title = "GDP per capita on Five continents") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
p + geom_line(aes(group = country), color = "gray70") +
  geom_smooth(size = 1.1, method = "loess", se = FALSE) +
  scale_y_log10(labels = scales::dollar) +
  facet_grid(~continent) + # for single categorical variable; for multiple categorical variables use facet_grid()
  labs(x = "Year",
       y = "GDP per capita",
       title = "GDP per capita on Five continents") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


## Transforming

### Use pipes to summarize data

Also, we experiment bar charts here. By default, `geom_bar` [uses](https://www.rdocumentation.org/packages/ggplot2/versions/1.0.1/topics/geom_bar) stat = "bins", which makes the height of each bar equal to the number of cases in each group. If you have a y column, then you should use `stat = "identity"` argument. Alternatively, you can use `geom_col()`.

```{r}

gapminder_formatted <- gapminder %>%
  group_by(continent, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp))

ggplot(data = gapminder_formatted, aes(x = year, y = lifeExp_mean, color = continent)) +
  geom_point() +
    labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy on Five continents")

gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = lifeExp_mean, color = country)) +
  geom_point() +
    labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in Europe")
```

```{r}
# geom point 
gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = lifeExp_mean)) +
  geom_point() +
    labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  facet_wrap(~country)

# geom bar
gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = lifeExp_mean)) +
  geom_bar(stat = "identity") + 
    labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  facet_wrap(~country)

# no facet 
gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = lifeExp_mean, fill = country)) +
  geom_bar(stat = "identity") + # even if you not stack, still the plot looks messy or you can use geom_col()
    labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in Europe")
```

```{r}

gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = country, y = lifeExp_mean)) +
  geom_boxplot() +
    labs(x = "Country",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  coord_flip()

```

```{r}
# without ordering
gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = reorder(country, lifeExp_mean), y = lifeExp_mean)) +
  geom_boxplot() +
    labs(x = "Country",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  coord_flip()

# reorder
gapminder %>%
  filter(continent == "Europe") %>%
  group_by(country, year) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = reorder(country, -lifeExp_mean), y = lifeExp_mean)) +
  geom_boxplot() +
    labs(x = "Country",
       y = "Life expectancy",
       title = "Life expectancy in Europe") +
  coord_flip()
```

### Plotting text

```{r}
gapminder %>%
  filter(continent == "Asia" | continent == "Americas") %>%
  group_by(continent, country) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = gdp_mean, y = lifeExp_mean)) +
  geom_point() +
  geom_text(aes(label = country)) +
  scale_x_log10() +
  facet_grid(~continent)
```

```{r}
# with label 
gapminder %>%
  filter(continent == "Asia" | continent == "Americas") %>%
  group_by(continent, country) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = gdp_mean, y = lifeExp_mean)) +
  geom_point() +
  geom_label(aes(label = country)) +
  scale_x_log10() +
  facet_grid(~continent)
```

```{r}
# no overlaps
gapminder %>%
  filter(continent == "Asia" | continent == "Americas") %>%
  group_by(continent, country) %>%
  summarize(gdp_mean = mean(gdpPercap),
            lifeExp_mean = mean(lifeExp)) %>%
  ggplot(aes(x = gdp_mean, y = lifeExp_mean)) +
  geom_point() +
  geom_text_repel(aes(label = country)) + # there's also geom_label_repel
  scale_x_log10() +
  facet_grid(~continent)
```

## Ploting models 

In plotting models, we extensively use David Robinson's [broom package](https://cran.r-project.org/web/packages/broom/vignettes/broom.html) in R. The idea is to transform model outputs (i.e., predictions and estimations) into tidy objects so that we can easily combine, separate, and visualize these elements. 

### Plotting several fits at the same time
```{r}
model_colors <- RColorBrewer::brewer.pal(3, "Set1") # select three qualitatively different colors from a larger palette.

gapminder %>%
  ggplot(aes( x= log(gdpPercap), y = lifeExp)) +
    geom_point(alpha = 0.2) +
    geom_smooth(method = "lm", aes(color = "OLS", fill = "OLS")) +
    geom_smooth(method = "lm", formula = y ~ splines::bs(x, df = 3),
                aes(color = "Cubic Spline", fill = "Cubic Spline")) +
    geom_smooth(method = "loess", aes(color = "LOESS", fill = "LOESS")) +
    theme(legend.position = "top") +
    scale_color_manual(name = "Models", values = model_colors) +
    scale_fill_manual(name = "Models", values = model_colors)
```

### Extracting model outcomes 
```{r}

# regression model
out <-lm(formula = lifeExp ~ gdpPercap + pop + continent, 
          data = gapminder)

```

Tidy is a method in broom package. It "constructs a dataframe that summarizes the model's statistical findings". As the description states, tidy is a function that can be used for various models. For instance, a tidy can extract following information from a regression model.

- Term: a term being estimated 
- p.value
- statistic: a test statistic used to compute p-value
- estimate 
- conf.low: the low end of a confidence interval 
- conf.high: the high end of a confidence interval
- df: degrees of freedom

**Challege**

Try glance(out), what did you get from these commands? If you're curious, you can try ?glance.

The followings are to show your degree of confidence.

#### Coeffficients
```{r}
# estimates 
out_comp <- tidy(out)

p <- out_comp %>%
  ggplot(aes(x = term, y = estimate))

p + geom_point() +
  coord_flip() +
  theme_bw()
```

#### Confidence intervals
```{r}
# plus confidence intervals 
out_conf <- tidy(out, conf.int = TRUE)

# plotting coefficients using ggplot2 (pointrange)
out_conf %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() + coord_flip() + labs(x = "", y = "OLS Estimate") +
  theme_bw()

# another way to do it (errorbar)
out_conf %>%
  ggplot(aes(x = estimate, y = reorder(term, estimate))) +
  geom_point() + 
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  labs(y = "", x = "OLS Estimate") +
  theme_bw()
```

You can calculate marginal effects using `margins` package. For the sake of time, I'm not covering that here.

**Challenge**

1. If we only want to visualize a certain subset of variables, let's say gdpPercap and pop, how can you do that? Also, gdpPercap might be not very informative. What's the best way to change the value name?

2. broom is a great package for running split-and-combine regressions. See the following example and write down your workflow for visualize it.


```{r}
gapminder %>%
  group_by(continent) %>%
    do(tidy(lm(gdpPercap ~ lifeExp, data = .), conf.int = TRUE))

?do
```

3. You can write a function for OLS and apply that function to grouped subsets of the data. I don't expect you to do this since we're covering functions in much detail next week.

```{r}
fit_ols <- function(df){
  lm(lifeExp ~ log(gdpPercap), data = df)
}
  
out_le <- gapminder %>%
  group_by(continent, year) %>%
  nest() %>% # nest() is an intensive version of group_by(). It creates a data column.
  mutate(model = map(data, fit_ols)) # the data column comes from nest().

gapminder %>%
  group_by(continent, year) %>%
  nest()
```

