---
title: 'More Advanced Programming in R: Lecture Notes'
author: "PS239T"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

We've already learned about types in R. Now let's discuss some more advanced programming in R - namely:

1. [String operations](#strings)
2. [Flow control](#flow-control)
3. [Functions](#functions)
4. [Loops](#loops)
5. [Vectorized Functions](#vectorization)

# Setup environment

```{r eval=F}
# remove all objects
rm(list=ls())

```

# 1. Strings {#strings}

### 1.1 Intro to Operations on Strings

```{r}
firstName <- "Jae Yeon"
lastName <- "Kim"
```

Unlike in Python, R does not have a reserved operator for string concatenation such as `+`.  Furthermore, using the usual concatenation operator ```c()``` on two or more character strings will not create a single character string, but rather a **vector** of character strings. 

```{r}
fullName <- c(firstName, lastName)

print(fullName)

length(fullName)
```

In order to combine two or more character strings into one larger character string, we use the ```paste()``` function.  This function takes character strings or vectors and collapses their values into a single character string, with each value separated by a character string selected by the user.

```{r}
fullName <- paste(firstName, lastName)

print(fullName)

fullName <- paste(firstName, lastName, sep = "+")

print(fullName)

fullName <- paste(firstName, lastName, sep = "___")
print(fullName)
```

As with Python, R can also extract substrings based on the index position of its characters.  There are, however, two critical differences.  First, **index positions in R start at 1**.  This is in contrast to Python, where indexation starts at 0.  

Second, **object subsets using index positions in R contain all the elements in the specified range**.  If some object called ```data``` contains five elements, ```data[2:4]``` will return the elements at the second, third, and fourth positions.  By contrast, the same subset in Python would return the objects at the third and fourth positions (or second and third positions, depending upon whether your index starts at 0 or 1).  

Third, **R does not allow indexing of character strings***. Instead, you must use the ```substr()``` function.  Note that this function must receive both the ```start``` and ```stop``` arguments.  So if you want to get all the characters between some index and the end of the string, you must make use of the ```nchar()``` function, which will tell you the length of a character string.

```{r}

fullName <- paste(firstName, lastName)

# this won't work like in Python
fullName[1] # R sees the string as a unitary object - it can't be indexed this way
fullName[1:4] 

# So use this instead
substr(x = fullName, start = 1, stop = 2)
substr(x = fullName, start = 5, stop = 5)
substr(x = fullName, start = 1, stop = 10)
substr(x = fullName, start = 11, stop = nchar(fullName))
```

### 1.2 Character String Methods

Like Python, R has a number of string methods, though these exist as individual rather than "mix-and-match" functions. For example:

```{r}
toupper(x = fullName)
tolower(x = fullName)

strsplit(x = fullName, split = " ")
strsplit(x = fullName, split = "n")

gsub(pattern = "Kim", replacement = "Choi", x = fullName)
gsub(pattern = "Jae Yeon", replacement = "Danny", x = fullName)

# Note the importance of cases! This doesn't throw an error, so you won't realize your function didn't work unless you double-check several entries 

gsub(pattern = " ", replacement = "", x = fullName) # The same function is used for replacements and stripping
```

# 2. Flow Control {#flow-control}

Almost all the conditional operators used in Python also work in R.  The basic loop set up is also very similar, with some small syntax adjustments.  Note that ```if()``` is a function whose arguments must be specified inside parentheses.  ```else```, however, is a reserved operator that takes no arguments.  Note that there is no ```elif``` option --- one simply writes ```else if()```.  Whereas operations to be executed after conditional evaluations in Python come after a ```:```, R operations must only be enclosed in curly brackets: ```{}```.  Furthermore, there is no requirement for indentation.  The only thing to keep in mind is that **each new operation must be on a separate line**.

```{r}
x <- 5

if(x < 0){
  print("x is negative")
} 

x <- -5

if(x < 0){
  print("x is negative")
}

x <- 5

if(x < 0){
  print("x is negative")
} else{
  print("x is positive")
}

x <- 0

if(x < 0){
  print("x is negative")
} else if(x == 0){
  print("x is zero")
} else{
  print("x is positive")
}

```

R also does some class coercion that makes Boolean evaluations harder to break than in Python.  But be careful --- R has a set of special coercion used for fast logical evaluation and subsetting.  Specifically, ```TRUE``` is considered equal to ```1```, while ```FALSE``` is equal to ```0```. The Boolean logicals can also be specified as a full word in all caps, or simply as ```T``` or ```F```.

```{r}
1 < 2
"1" < 2
"a" < 2

TRUE < 2

TRUE == "TRUE"
T == "TRUE"
TRUE == "T"
TRUE == "FALSE"

TRUE == 0
TRUE == 1
FALSE == 0
FALSE <= 1
```

# 3. Functions {#functions}

While functions are defined in Python using the ```def``` reserved operator, R sees functions as just another type of named object.  Thus, they require explicit assignment to an object.  This is done using the function ```function()```, which creates a function taking the arguments specified in parentheses.  

```{r, eval = FALSE}
df$a[df$a == -99] <- NA
df$b[df$b == -99] <- NA
df$c[df$c == -99] <- NA
df$d[df$d == -99] <- NA
df$e[df$e == -99] <- NA
df$f[df$f == -99] <- NA 

# very tedious 

fix_missing <- function(x) {
  x[x == -99] <- NA
  x
}

df$a <- fix_missing(df$a)
df$b <- fix_missing(df$b)
df$c <- fix_missing(df$c)
df$d <- fix_missing(df$d)
df$e <- fix_missing(df$e)
df$f <- fix_missing(df$e)

# beautiful

source(fix_missings.r)

# more beautiful
```

```{r}
simple.function <- function(x){
  print(x + 1)
}

simple.function(x = 2)

less.simple.function <- function(x, y){
  print(x - y + 1)
}

less.simple.function(x = 2, y = 10)
```

With respect to returning function output, most of the same rules apply as with Python. Be sure to remember that ```return()``` will only process a single object, so multiple items must usually be returned as a list. Note that your ordering of the functions matters, too. 

```{r}
dumbfun <- function(x){
  return(x)
  print("This will never print :(")
}

dumbfun(x = "something")

dumbfun <- function(x){
  print("Why did I print?")
  return(x)
}

dumbfun(x = "something")

dumbfun <- function(x,y){
  thing1 <- x
  thing2 <- y
  return(list(thing1, thing2))
}

dumbfun(x = "some text", y = "some data")
dumbfun(x = c(5,10,15), y = "some data")
```

R functions also allow you to set default argument values:

```{r}
less.simple.function <- function(x, y = 0){
  print(x - y + 1)
}

less.simple.function(x = 2)

less.simple.function(x = 2, y = 10)
```

With respect to specifying arguments, one can either use argument **position** specifications (i.e., the order) or argument **name** specifications.  The latter is strongly preferred, as it is very easy to accidentally specify incorrect argument values.

```{r}
send <- function(message, recipient, cc=NULL, bcc=NULL){
  print(paste(message, recipient, sep = ", "))
  print(paste("CC:", cc, sep = " "))
  print(paste("BCC:", bcc, sep = " "))
}

send(message = "Hello", recipient = "World", cc = "Sun", bcc = "Jane")

send("Hello", "World", "Sun", "Jane")

send("Hello", "Sun", "Jane", "World")

send(message = "Hello", cc = "Sun", bcc = c("Jane", "Rochelle"), recipient = "World")
```

Also, note that functions don't have what CS people called side-effects. Functions only define local variables = They don't change objects stored in the global environment. (Consider the difference between `<-` and `=` for assignments.) That's why you can use functions for reusable tasks since it does not interrupt other important things in your system.

See [the following example](https://darrenjw.wordpress.com/2011/11/23/lexical-scope-and-function-closures-in-r/) from Wilkinson.

```{r}
a = 1 
b = 2

f <- function(x)
{
  a*x + b
}

f(2)

g <- function(x)
{
  a = 2
  b = 1
  f(x)
}

g(2) # a equals still 1 
```

# 4. Loops {#loops}

Loops in R also work basically the same way as in Python, with just a few adjustments.  First, recall that index positions in R start at 1.  Second, ```while()``` and ```for()``` are functions rather than reserved operators, meaning they must take arguments in parentheses.  Third, just like ```else```, the ```in``` operator *is* reserved and takes no arguments in parentheses.  Fourth, the conditional execution must appear between curly brackets.  Finally, indentation is meaningless, but each new operation must appear on a new line.

- `while()`: when we have no idea how many times loop needs to be executed.
- `for()`: when we know how many times loop needs to be executed.

```{r}
fruits <- c("apples", "oranges", "pears", "bananas")

# a while loop
i <- 1
while(i <= length(fruits)){
  print(fruits[i])
  i <- i + 1
}

# a for loop
for(i in 1:length(fruits)){
  print(fruits[i])
}
```

# 5. Vectorization {#vectorization}

### 5.1 Vectorized functions

Most of R’s functions are vectorized, meaning that the function will operate on all elements of a vector without needing to loop through and act on each element one at a time. This makes writing code more concise, easy to read, and less error prone.

```{r}
x <- 1:4
x * 2
```

The multiplication happened to each element of the vector, simultaneously.

We can also add two vectors together:

```{r}
x <- 1:4
y <- 6:9
x + y
```

Each element of x was added to its corresponding element of y:

```{r eval = F}
x:  1  2  3  4
    +  +  +  +
y:  6  7  8  9
---------------
    7  9 11 13
```

Most functions also operate element-wise on vectors:

```{r}
x <- 1:4
log(x)
```

To learn more about using and speeding up for loops, check out the answers to this post: http://stackoverflow.com/questions/2908822/speed-up-the-loop-operation-in-r

### 5.2 Apply Family

While and for loops in R can be very slow. For this reason, R has a number of built-in iteration methods to speed up execution times. In many cases, packages will have "behind-the-scenes" ways to avoid for loops, but what if you need to write your own function? 

A common method of getting around for loops is the **apply** family of functions. These take a data structure and a function, and applies a function over all the elements in the object.

```{r}
fruit <- c("apple", "orange", "pear", "banana")

# make function that takes in only one element
make.plural <- function(x){
   plural <- paste(x, 's', sep = '') # sep is for collapse, so collpase ''
   return(plural)
}

make.plural('apple')

# apply that function to every element
lapply(fruit, make.plural) # returns a list
sapply(fruit, make.plural) # returns a named vector

library(purrr) # load package
map(fruit, make.plural) # type consistent
```

```{r}
# Why sapply is bad 

sapply(1:100, paste) # return character 

sapply(integer(), paste) # return list!

library(purrr)
map(1:100, paste) # return list

map(integer(), paste) # return list
```

The multivariate version of `sapply` is `mapply`. Use this if you have a function that takes in 2 or more arguments.

# 6. Debugging 

## 6.1. RStudio debugging tools 

Debugging is a process. Don't use this process when you're writing a new function, but when you try to fix it. The idea is try to fail systematically.

- Realize that you have a bug.
- Make it repeatable. (Remove the code until you get to the smallest snippet that still causes the error.)
- Locate the bug.
- Fix it and test it.

RStudio has several debugging tools. 
- `traceback`: list the sequence of calls that lead to the error.
- Rerun with Debug tool and `options(error = browser)`: opens an interactive session where the error occurred.
- breakpoints and `browser()`: opens an interactive session at an arbitrary location in the code.

```{r, eval = FALSE}
f <- function(a) g(a)
g <- function(b) h(b)
h <- function(c) i(c)
i <- function(d) "a" + d #"a" and d are non-numeric and + is a binary operator

f(10)
```

## 6.2. Condition handling 

- `try()`: continue execution even when an error occurs.
- `tryCatch()`: specify handler functions when a condition is signalled.

```{r, eval = FALSE}
f1 <- function(x) {
  log(x)
  10
}

f1("x")

f2 <- function(x) {
  try(log(x))
  10
}

f2("a") # Now it's running through with an error

```

```{r}
show_condition <- function(code) {
  tryCatch(code,
    error = function(c) "error",
    warning = function(c) "warning",
    message = function(c) "message"
  )
}

show_condition(stop("!")) # error
show_condition(warning("?!")) # warning
show_condition(message("?")) # message 

```

** Fail fast (defensive programming) **

- Be strict about what you accept. You can use `stopifnot` from `assertthat` package, or simple `if` statements and `stop()`.
- Avoid functions that use [non-standard evaluations](http://adv-r.had.co.nz/Computing-on-the-language.html#nse) like subset. When they fail, they often fail with uninformative error messages.
- Avoid functions that return different types of output depending on input like `sapply` (which returns different types of vectors depending on input types). Never use, `sapply` inside a function. 
- `map` function from purrr package is a better alternative if you take type-consistency seriously.
```