---
title: "Style Guide"
output:
  pdf_document: default
  html_notebook: default
author: "PS239T"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r}
# Most of contents here are adapted from https://style.tidyverse.org 
```


# Five rules

- Write code as if you're writing a paper.

- Organize files. 

- Comment religiously. 

- Reuse functions. 

- Do version control. 


# Files 

```{r, eval= FALSE}
# Good
fit_models.R

# Bad
fit models.R
```

- When naming files:
    - Don't use special characters.
    - Don't capitalize.
    - Numbering them if files should be run in an order.
    
# Syntax 

## Objects 

```{r, eval = FALSE}
# Good 
day_one
    
# Bad 
DayOne
```

- When naming objects:
    - Don't use special characters.
    - Don't capitalize.
    - Numbering them if files should be run in an order.

## Functions 

```{r, eval = FALSE}
# Good 
run_rdd 

# Bad 
rdd
```

- When naming objects:
    - Don't use special characters.
    - Don't capitalize.
    - Use verbs instead of nouns.
    
## Spacing 

```{r, eval = FALSE}
# Good
x[, 1] 

mean(x, na.rm = TRUE) 

if (debug) show(x) 

function(x) {}

height <- (feet * 12) + inches

sqrt(x^2 + y^2)

# Bad
x[,1]

mean (x, na.rm = TRUE)

if(debug) show(x)

function(x){}

height <- (feet*12) + inches

sqrt(x ^ 2 + y ^ 2)
```

## Indenting 

```{r eval= FALSE}
# Good
if (y < 0 && debug) {
  message("y is negative")
}

# Bad
if (y < 0 && debug)
message("Y is negative")
```

## Long lines

```{r, eval = FALSE}
# Good
do_something_very_complicated(
  something = "that",
  requires = many,
  arguments = "some of which may be long"
)

# Bad
do_something_very_complicated("that", requires, many, arguments,
                              "some of which may be long"
                              )
```

## Comments 

- Use comments for findings and analysis decisions. 
- Show your code; Do not try to explain your code by comments.

# Functions 

```{r, eval = FALSE}
# Good
add_row()

# Bad
row_adder()
```

## Comments 

- Inside functions, use comments to explain **why**

# Pipes 

```{r, eval = FALSE}
# Good
iris %>%
  group_by(Species) %>%
  summarize_if(is.numeric, mean) %>%
  ungroup() %>%
  gather(measure, value, -Species) %>%
  arrange(value)

# Bad
iris %>% group_by(Species) %>% summarize_all(mean) %>%
ungroup %>% gather(measure, value, -Species) %>%
arrange(value)

```

# Additional tips 

- Use styler package in R to learn how to format your code to according to the style guide above

