---
title: "Dictionary methods in R"
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

The following material is adapted from [Text Mining with R](https://proquest.safaribooksonline.com/9781491981641) (2017) by Julia Silge and David Robinson.

# Setup
```{r}

rm(list=ls())

# load packages
require(tm) # for text mining
library(tidyverse) # for tidyverse
library(tidytext) # for tidy text analysis
library(quanteda) # for text analysis and examples
library(wordcloud) # for wordcloud
library(stringr) # for string manipulation
library(reshape2)

# load data 
inaug <- tidy(data_corpus_inaugural)
```

# Sentiments 

- These sentiment lexicons are from [AFINN](http://www2.imm.dtu.dk/pubdb/views/publication_details.php?id=6010) from [Finn Arup Nielson](http://people.compute.dtu.dk/faan/), [Bing](https://www.cs.uic.edu/~liub/FBS/sentiment-analysis.html) from [Bing Liu](https://www.cs.uic.edu/~liub/) and collaborators, and [NRC](http://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm) from [Saif Mohammad](http://saifmohammad.com/) and [Peter Turney](https://www.apperceptual.com/).

- NRC lexicons, and the Bing lexicons are measured using categorical variables and the AFINN lexicons are based on a scale of -5 to 5.

- These lexicons are validated by either crowdsourcing or the authors or both of them. Basically, this is a case of a dictionary method in which a dictionary is about sentiments.

```{r}
# example
sentiments %>%
  group_by(sentiment) %>%
  filter(sentiment == "trust")

# call a sentiment lexicon 
get_sentiments("nrc")
```

# Sentiment analysis

## Plot

```{r}

# apply to the data 


## AFINN
inaug %>%
  filter(Year > 1945) %>%
  unnest_tokens(word, text) %>%
  inner_join(get_sentiments("afinn")) %>%
  count(President, Year, score) %>%
  ggplot(aes(x = score, y = n, col)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(President ~ Year, scale = "free_y") +
  coord_flip()

## NRC
inaug %>%
  filter(Year > 1945) %>%
  unnest_tokens(word, text) %>%
  inner_join(get_sentiments("nrc")) %>%
  count(President, Year, sentiment) %>%
  ggplot(aes(x = sentiment, y = n, col)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(President ~ Year, scale = "free_y") +
  coord_flip()

## Bing
inaug %>%
  filter(Year > 1945) %>%
  unnest_tokens(word, text) %>%
  inner_join(get_sentiments("bing")) %>%
  count(President, Year, sentiment) %>%
  ggplot(aes(x = sentiment, y = n, col)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(President ~ Year, scale = "free_y") +
  coord_flip()
```

## Count

- What's going on here? Let's count most frequent words associated with positive and negative sentiments using the Bing lexicons.

```{r}
## Bing
inaug %>%
  filter(Year > 1945) %>%
  unnest_tokens(word, text) %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE)
```

## Wordcloud

- Frankly, not a big fan of this visualization method. But can be useful for the data exploration.

```{r}

# without sentiment analysis
inaug %>%
  filter(Year > 1945) %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words) %>%
  count(word) %>%
  with(wordcloud(word, n, max.words = 100))

# with sentiment analysis
inaug %>%
  filter(Year > 1945) %>%
  unnest_tokens(word, text) %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  reshape2::acast(word ~ sentiment, value.var = "n", fill = 0) %>% # turn the dataframe into a matrix
  comparison.cloud(colors = c("red", "blue"),
                   max.words = 100)

```

## Creating your own dictionary 

```{r}
# create your own dictionary
race_dictionary <- data.frame(word = c("african", "negro", "black", "asian", "oriental", "latino", "hispanic", "nonwhite"), category = c("Afircan Americans", "African Americans", "African Americans", "Asian Americans", "Asian Americans", "Latinx", "Latinx", "People of Color"))

class_dictionary <- data.frame(word = c("inequality","poverty","growth"), category = c("Distribution","Distribution","Growth"))
# dictionary-based analysis
  
## race
inaug %>%
  unnest_tokens(word, text) %>%
  inner_join(race_dictionary) %>%
  count(President, Year, category) %>%
  ggplot(aes(x = category, y = n, col)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(President ~ Year, scale = "free_y") +
  coord_flip()
  
## class
inaug %>%
  unnest_tokens(word, text) %>%
  inner_join(class_dictionary) %>%
  count(President, Year, category) %>%
  ggplot(aes(x = category, y = n, col)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(President ~ Year, scale = "free_y") +
  coord_flip()
```



